A Recipe for Scavenging - the natural history of a behaviour

### 2016-04-07 natural history of scavenging.tex
More edits!

%\affiliations\{
%\item A. Kane (\url{adam.kane@ucc.ie}), University College Cork, Cooperage Building, School of Biological Earth and Environmental Sciences, Cork, Ireland.
%\item K. Healy and A. Jackson, Trinity College Dublin, Department of Zoology, School of Natural Sciences, Dublin Ireland.
%\item T. Guillerme, Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.
%\item G. Ruxton, School of Biology, Sir Harold Mitchell Building, Greenside Place, St Andrews, KY16 9TH, United Kingdom.
%}

@@

% \maketitle

@@

\subsection*{The Difficulty of Scavenging}
The chief hurdle of scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some level of opportunistic scavenging chances. %TG: Two times "opportunities", replaced by "chances"
However, species that rely on scavenging to sustain substantial portions of their diets must increase the probability of encountering a sufficient amount of carrion in order to meet their energetic demands.
%Once found, the scavenger must be able to out-compete any potential competitors and process a product under the processes of decay from toxin producing and disease causing microorganisms \citep{ruxton2014fruit}. %TG: make the following sentence more sense?
Once found, the scavenger must be able to out-compete any potential competitors and process the carrion that is undergoing decay and thus contains toxins and diseases produced by micro-organisms \citep{ruxton2014fruit}.
Finally, the level of scavenging opportunities available will also depend on the density, size, and quality of carcases produced, all of which are dependent on complex ecosystem dynamics.
These facets which are required for scavenging are essentially the key parameters found in the species functional response curves of optimal foraging theory, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. %TG: Jaysus, this sentence doesn't read well! "Functional response curves of optimal foraging theory" can probably be phrased in a nicer/easier way...
We compare the ability of scavenging across vertebrates focusing on each of these facets and their relationship with scavenging.
By considering scavenging in a framework of optimal foraging and identifying the prerequisite attributes and processes required for scavenging we propose a ``scale of scavenging'' whereupon we can place any vertebrate species, past or present, and assess the importance of scavenging in its diet.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between the consumer and resources. 
In the simplest case, this rate can be thought of in terms of a gas diffusion model where the movement of two agents depends only on their relative speed.
As carcasses are stationary, the relative speed between a scavenger and carrion is only dependent on the movement of the scavenger. 
Vertebrates do more than simply bump into resources like gas molecules though, because they can actively detect them through their sensory abilities. 
Therefore, scavenging will also depend on search rates, which are dependent on both species physiology and the dimensionality of the environment \citep{pawar2012dimensionality}.

\subsection*{Locomotion}
In contrast to the importance of locomotion in predation, locomotion relating to scavenging depends more on the ability to efficiently move over large areas. 
This ability generally requires an efficient transfer of metabolic energy into movement which relies on both physiology (i.e. metabolism) and the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is, paradoxically, found in flying species. 
Despite the energetic costs of flight, the only known vertebrate obligate scavengers are the old and the new world vultures. 
This is because, although powered flight is energetically expensive, species like vultures can exploit air currents using large wingspans which allows them to soar while only doubling their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}. %TG: I didn't get the meaning of this sentence, is soaring doubling the metabolic rate? or is it a word misplacement ("while only doubling their metabolic rate")? and if so, how does that make the point?
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
This mode of locomotion is also exploited by sea birds, which use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} % KH and are likley to have being soaring fliers as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?).  AK I don't think this is true from what I've seen
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
Thus, the use of soaring in extinct species as a means for scavenging is also likely.

While soaring is perhaps the only viable means of locomotion that allows for an obligate, scavenging life-style \citep{ruxton2004obligate} powered flight is still an efficient means of locomotion. 
Avian flight is a cheaper means of locomotion than either walking or running \citep{tucker1975energetic} 
Even taking account of maintenance costs this still bears out, where total cost of movement (J kg\textsuperscript{-1} m\textsuperscript{-1}) scales according to 5.2 $\times$ body mass (kg) \textsuperscript{-0.23} for fliers and 10.7 $\times$ body mass (kg) \textsuperscript{-0.32} for runners \citep{williams1999evolution}.

@@

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the total cost of movement (again in J kg\textsuperscript{-1} m\textsuperscript{-1}) in fish is lower than either running or flying where it scales according to 2.15 $\times$ body mass (kg) \textsuperscript{-0.25} \citep{williams1999evolution} with only \textit{soaring} flight likely to surpass it. 
%KH(I assume the reference is for powered flight, again this needs to be more precise).  
This has led some researchers to argue for the likelihood of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
Sharks perhaps best resemble the large soaring fliers as they depend on large pectoral fins in order to maintain lift as they swim. 
Many shark species show very large ranges (e.g. great white sharks (\textit{Carcharodon carcharias})) and have the lowest cost of swimming.%TG: I guess the following is a comment: (I know the Greenland shark is the slowest swimmer). 
Facultative scavenging is seen in many marine groups including species of extant sharks such as great white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}.
There is evidence too of scavening in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and mysticete whales from the Pliocene \citep[5.3 to 3.6 Mya;]{schwimmer1997scavenging,ehret2009caught}.

@@

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya %TG: that early???
have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system.

%\cite{williams1999evolution} notes, ``free-ranging animals must contend with the total energetic expenditure associated with supporting basic biological functions as well as with moving the body and appendages through the environment."

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, because an animal must trade off the cost of supporting its body off the ground while propelling itself forward. 
The idea that the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals conferred a huge advantage to the latter is often mentioned \citep[e.g. benefits in terms of speed, efficiency, muscle effort and manoeuvrability; ]{sullivan2015posture}. 
Yet, as \cite{sullivan2015posture} states, most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Although, metabolic rate unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also be true of extinct species. 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

Modern endothermic mammals can sustain longer periods of energetically expensive activity \citep{bennett1979endothermy} resulting in larger foraging ranges. 
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass.
In the case of mammals and reptiles these are 1.15 $\times$ body mass (kg) \textsuperscript{0.12} and 0.23 $\times$ body mass (kg) \textsuperscript{0.12} respectively \citep{ruxton2004obligate}.
If we insert these into a foraging radius model $\sfrac{\frac{\text{duration} \times \text{speed}}{2}}{1000}$ for a 12 hour foraging day which shows that while a 10 kg reptile can range 6.5 km, an equally sized mammal can range nearly 33 km \citep{Enstipp2006Energetics}. 
For a foraging scavenger, this ability translates into a greater area searched for food.

@@

Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures %TG: We need a standardisation for english species names, I'll go for all lower case (similar as what you'd do for "pig" or other english common species names).
 (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed visual abilities \citep{reymond1985spatial}.
It follows from this that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.
%We can contrast this ability with bats, whose visual acuity is famously poor. %TG: Humph! That's not nice! I'm wouldn't go with all the "this is known" thingies here. It really lowers down the quality of our claim and is totally un-necessary (and actually probably wrong: lot's of fruit bats (like flying foxes) do not echolocate and thus have to find their fruits among the flipping tropical folliage using their eyes and noses only)...
We can contrast that with bats for which echolocation would not lend itself to discover immobile carrion.
Additionally, their small size and poor terrestrial ability would also count against them when feeding at a carcass \citep{riskin2006terrestrial}.

@@

Having a panoramic view means being able to gather a wealth of information from other foragers, be they conspecifics or otherwise \citep{jackson2008effect}.
Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}.
These behaviours allow them forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \citep{KaneVul}.

@@

\section*{Handling Time}
Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over.
Being able to extract nutrients from remnants gives a scavenger a great advantage.
%Thus, the bone crushing ability of hyenas and others reveals another useful scavenger trait. %TG: how about the following for a smoother transition:
Thus, some phenotypic adaptation such as the ability to crush bone (like in hyenas) reveals another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have been compared to hyenas in terms of their feeding ecology as well \citep{van2003chapter,martin2016pursuit}.

% Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology.
% As a result, the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.
% And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores.
% Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past.
% These traits have been used to infer scavenging lifestyles in extinct mammals. 
%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.
Certainly, scavenging should be particularly attractive to avian predators compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.
%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.

%TG: This part above makes sense in the adaptations parts (i.e. in the Handling Time part) but we can see when we clean down the whole paper if we keep it (then we need to integrate it) or if we have enough. I actually fits way better in the new narrative by the way!

@@

Large body size confers substantial dominance and starvation-resistance benefits \citep{ruxton2004obligate}.
Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers.
Wandering Albatrosses (\textit{Diomedea exulans}), cinereous vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{weimerskirch1992reproductive,ferguson2001raptors,donazar2002effects}.

And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of over 200 kg \citep{witton2010size}.
Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against pterosaurs existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficiency indicates they would have been comfortable foraging on the ground.
Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) have a comparable morphology and are noted facultative scavengers \citep{monadjem2012survival} so it is reasonable to believe that certain pterosaurs behaved similarly.

The competitive ability of even the largest living bird is radically diminished in their interactions with mammalian competitors however, and as such they tend to consume carrion rapidly. %TG: I've added "living" bird here. In the Paleocene you have these giant terror birds that were still top predators (like during the Cretaceous... How weird! ;)).
\cite{houston1974role} observed a group of \textit{Gyps} vultures consuming all of the soft tissue from a 50 kg Grant’s gazelle (\textit{Nanger granti}) in eight minutes. 
Their serrated tongues and hooked bills enabling them to achieve this feat \citep{houston1975digestive}. 
Outside of raptors such as vultures, the specialised beaks of many modern bird lineages hinders their ability to eat meat which is in contrast to the first lineages that did not have this feature \citep{martyniuk2012field} . 
As \cite{martyniuk2012field} notes these early birds would thus have been predominantly carnivorous, which implies that scavenging would have been a live opportunity cf. their descendants.

>>>>

\affiliations{
\item A. Kane (\url{adam.kane@ucc.ie}), University College Cork, Cooperage Building, School of Biological Earth and Environmental Sciences, Cork, Ireland.
\item K. Healy and A. Jackson, Trinity College Dublin, Department of Zoology, School of Natural Sciences, Dublin Ireland.
\item T. Guillerme, Imperial College London, Silwood Park Campus, Department of Life Sciences, Buckhurst Road, Ascot SL5 7PY, United Kingdom.
\item G. Ruxton, School of Biology, Sir Harold Mitchell Building, Greenside Place, St Andrews, KY16 9TH, United Kingdom.
}

@@

 \maketitle

@@

\subsection*{The Difficulty of Scavenging}
The chief hurdle of scavenging is finding a resource that is often difficult to predict in space and time.
Through chance alone many species will avail of some opportunistic scavenging. %TG: Two times "opportunities", replaced by "chances"
However, species that rely on scavenging to sustain substantial portions of their diets must increase the probability of encountering a sufficient amount of carrion in order to meet their energetic demands.
%Once found, the scavenger must be able to out-compete any potential competitors and process a product under the processes of decay from toxin producing and disease causing microorganisms \citep{ruxton2014fruit}. %TG: make the following sentence more sense?
Once found, the scavenger must be able to out-compete any potential competitors and process the typically decaying carcass which contains microorganism derived toxins  \citep{ruxton2014fruit}.
Finally, the potential for scavenging will also depend on the density, size, and quality of carcases produced, all of which are affected by complex ecosystem dynamics.
All of these facets are essentially the key parameters found in functional response curves, namely encounter rate, handling time and prey availability \citep{jeschke2002predator}. %TG: Jaysus, this sentence doesn't read well! "Functional response curves of optimal foraging theory" can probably be phrased in a nicer/easier way...
%With this framework we compare the ability of scavenging across vertebrates. % focusing on each of these facets and their relationship with scavenging.
By considering scavenging in this context of optimal foraging we can identify the prerequisite attributes and processes required for the behaviour. 
This has enabled us to propose a scale of scavenging whereupon we can place any vertebrate species, past or present, and assess the importance of carrion in its diet.

@@

%KH: need paragraph for how do we know something is (was) a scavenger
\section*{Encounter Rate}
All foraging processes depend on the encounter rate between the consumer and resources. 
In the simplest case, this rate can be thought of in terms of a gas diffusion model where the movement of two agents depends only on their relative speed.
As carcasses are stationary, the relative speed between a scavenger and carrion is only dependent on the movement of the scavenger. 
Vertebrates do more than simply bump into resources like gas molecules though, because they can actively detect them through their sensory abilities. 
As such, scavenging potential is strongly affected by search rates, which are determined by both species' physiology and the dimensionality of the environment \citep{pawar2012dimensionality}.

\subsection*{Locomotion}
Scavenging depends more on the ability to efficiently move over large areas than does predation because of the inherent unpredictability of carrion. 
This generally requires an efficient transfer of metabolic energy into movement which relies on both physiology (i.e. metabolism) and the medium of the environment in which the animal is moving (i.e. aerial, aquatic or terrestrial).
Perhaps the most efficient form of locomotion in vertebrates is, paradoxically, found in flying species. 
Despite the energetic costs of flight, the only known vertebrate obligate scavengers are the old and the new world vultures. 
And, although powered flight is energetically expensive, species like vultures can exploit air currents using their large wingspans which allows them to soar at a cost of only twice their metabolic rate \citep{hedenstrom1993migration,spivey2014analysing}. %TG: I didn't get the meaning of this sentence, is soaring doubling the metabolic rate? or is it a word misplacement ("while only doubling their metabolic rate")? and if so, how does that make the point?
By depending on thermal air flows these species can forage over vast ranges \citep{spiegel2013factors}. 
An analagous mode of locomotion is also exploited by sea birds, which use strong ocean winds to search large areas of the oceans \citep{norberg2012vertebrate,thaxter2012seabird}. 
While many species of seabird are likely primarily predators, it seems that albatrosses, who can range many hundreds of kilometres, take a substantial amount of carrion in their diet \citep{croxall1994dead}. 
This is typically in the form of squid, which float on the surface, allowing the birds to readily pluck their remains out of the water \citep{croxall1994dead}.

The two families from which modern vultures arose, the Accipitridae and Cathartidae, appear during the Palaeocene \citep[66 - 56 Million years ago (Mya); ][]{Jetz2012, Jarvis2014} % KH and are likley to have being soaring fliers as demonstrated by their large wingspans (Is this true?). (Is there any other form of soaring?).  AK I don't think this is true from what I've seen
However, soaring flight is likely to be far older than this with avian flight originating in the Late Jurassic (163.5-145 Mya) and vertebrate flight in the Late Triassic (235-201.3 Mya) coincident with the pterosaurs. 
Indeed, scavenging among pterosaurs has been hypothesised many times before \citep{witton2008reappraisal}. 
Certain groups of these animals could reach enormous sizes \citep[e.g. Azhdarchids with wingspans of 11 metres; ][]{witton2010size} and, notably, appear to have engaged in soaring flight \citep{witton2010size}.
It seems that the use of soaring in extinct species as a means for scavenging is likely.

While soaring is perhaps the only viable means of locomotion that allows for an obligate, scavenging life-style \citep{ruxton2004obligate}, powered flight is still an efficient means of locomotion. 
Avian flight is cheaper than either walking or running \citep{tucker1975energetic}. 
Even taking account of maintenance costs, this still bears out, where total cost of movement (J kg\textsuperscript{-1} m\textsuperscript{-1}) scales according to 5.2 $\times$ body mass (kg) \textsuperscript{-0.23} for fliers and 10.7 $\times$ body mass (kg) \textsuperscript{-0.32} for runners \citep{williams1999evolution}.

@@

Similar to aerial species, aquatic scavengers have a locomotory benefit because water is a medium that is conducive to low-cost movement \citep{tucker1975energetic}. 
In fact, the total cost of movement (again in J kg\textsuperscript{-1} m\textsuperscript{-1}) in fish is lower than either running or flying where it scales according to 2.15 $\times$ body mass (kg) \textsuperscript{-0.25} \citep{williams1999evolution} with only \textit{soaring} flight likely to surpass it. 
%KH(I assume the reference is for powered flight, again this needs to be more precise).  
This has led some researchers to argue for the likelihood of an obligate scavenging fish \citep{ruxton2004energetic,ruxton2005searching}. 
Sharks perhaps best resemble the large soaring fliers as they depend on large pectoral fins in order to maintain lift as they swim. 
Many shark species show very large ranges (e.g. great white sharks (\textit{Carcharodon carcharias})) and have the lowest cost of swimming. %TG: I guess the following is a comment: (I know the Greenland shark is the slowest swimmer). 
Facultative scavenging is seen in many marine groups including species of extant sharks such as great white sharks \citep[known to feed on whale carcasses;][]{fallows2013white}, Greenland sharks \citep[feeding on seals;][]{watanabe2012slowest}, and sixgill sharks \citep{anderson2016impact}.
There is evidence too of scavening in extinct species, where shark teeth have been found in the remains of dinosaurs, mosasaurs and mysticete whales from the Pliocene \citep[5.3 to 3.6 Mya;]{schwimmer1997scavenging,ehret2009caught}.

@@

% KH endotermy - why it doesn't seem to work
We might expect then that by combining an aquatic environment and an endothermic metabolism marine mammals would especially prosper as scavengers.
Fossil pinnipeds and cetaceans from 60 Mya %TG: that early???
have transitional features indicative of their trajectory to fully aquatic species \citep{williams1999evolution}.  
But despite their movement away from land their energetic savings were negligible because the \textit{total} cost incurred by a swimming marine mammal is high \citep{williams1999evolution}. 
Indeed, the total energetic cost is similar to an equivalent terrestrial or aerial mammal \citep{williams1999evolution}.
This underscores the trade offs between the benefits of endothermy in terms of activity periods and the costs of maintaining such an expensive system. 
That said, aquatic endotherms have and do scavenge. 
For instance, early whales such as \textit{Basilosaurus} seem to fit into the same niche as killer whales \textit{Orcinus orca} and we have some evidence for scavenging in both \citep{fahlke2012bite,Whitehead415}.
%\cite{williams1999evolution} notes, ``free-ranging animals must contend with the total energetic expenditure associated with supporting basic biological functions as well as with moving the body and appendages through the environment."

%% AK:  Need to be careful here because this whole idea of sprawling posture being worse off than erect posture is not that well supported in the literature. 
%% Great Transformations in Vertebrate Evolution By Kenneth P. Dial, Neil Shubin, Elizabeth L. Brainerd talks about it at length 
%% So I think this should stay pitched as an endo/ ectothermy thing rather than posture 
Terrestrial environments are the most energetically costly in which to move, because an animal must trade off the cost of supporting its body off the ground while propelling itself forward. 
The idea that the evolutionary transition in posture from the sprawling gait of reptiles to the erect posture of mammals conferred a huge advantage to the latter is often mentioned.
The purported advantages include benefits in terms of speed, efficiency, muscle effort and manoeuvrability \citep{sullivan2015posture}. 
Yet, as \cite{sullivan2015posture} states, most of the hypotheses in favour of this idea remain to be tested in the context of archosaur evolution. 
Metabolic rate however, unquestionably impacts terrestrial species whereby ectotherms such as many modern reptiles, cannot move for sustained periods \citep{bennett1979endothermy}. 
% Species which do not support an upright position such as modern reptiles limit their ability to move for sustained periods \citep{bennett1979endothermy}. 
This is exacerbated by their sprawling gait which results in the phenomenon known as Carrier's constraint such that the animal can't move and breathe at the same time because the lateral movements impedes its lungs \citep{carrier1987evolution}. 
% These constraints are likely to mean that many ectotermic terrestrial species, which cannot maintain an upright posture would depend on a relatively low level of scavenging (I'm not sure this is true though). 
This would also be true of extinct species. 
It is during the Permian (298.9 - 252.17 Mya) that we have the earliest evidence of vertebrate scavenging where a temnospondyl amphibian fed on the carcass of \textit{Varanops}, a predatory synapsid of the time \citep{reisz2006articulated}. 
%TG: I guess this is a comment too: Snakes are a bit different are more effecent at moving I think - lots of scavanging. 
Both these groups are more likely to depend on low energetic costs in order to compensate for low encounter rates due to their inability to move. 
It is with the evolution of endothermy in the therapsid-mammal lineage \citep{clarke2010temperature} that terrestrial vertebrates would have gained the ability to range more widely, a vital component in seeking out carrion.

Modern endothermic mammals can sustain longer periods of energetically expensive activity \citep{bennett1979endothermy} resulting in larger foraging ranges. 
To quantify this effect with a simple example we can turn to some allometric relationships relating sustainable travelling speed to body mass \citep{ruxton2004obligate}.
If we insert these into a foraging radius model \citep{Enstipp2006Energetics} for a 12 hour foraging day it shows that while a 10 kg reptile can range 6.5 km, an equally sized mammal can range nearly 33 km . 
For a foraging scavenger, this ability translates into a greater area searched for food.

@@

Species capable of flight have effectively added an extra spatial dimension, i.e. the vertical component, to their sensory environment over land animals.
This allows them to look down on a landscape where they are unencumbered by obstacles that would obstruct the view of a terrestrial scavenger.
Such an ability has obvious benefits in detecting carrion.
Certainly, vultures are known to have impressive visual acuity, with one estimate indicating lappet-faced vultures %TG: We need a standardisation for english species names, I'll go for all lower case (similar as what you'd do for "pig" or other english common species names).
 (\textit{Torgos tracheliotus}) are capable of detecting a 2 metre carcass over 10 km away \citep{spiegel2013factors}.
Eagles too are known to have highly developed vision \citep{reymond1985spatial}.
It follows that the evolution of flight allowed aerial animals to detect far more carrion than their terrestrial counterparts \citep{AR:AR22815}.
%We can contrast this ability with bats, whose visual acuity is famously poor. %TG: Humph! That's not nice! I'm wouldn't go with all the "this is known" thingies here. It really lowers down the quality of our claim and is totally un-necessary (and actually probably wrong: lot's of fruit bats (like flying foxes) do not echolocate and thus have to find their fruits among the flipping tropical folliage using their eyes and noses only)...
We can contrast this with many bats whose reliance on echolocation would not lend itself to discovering immobile carrion.
%

@@

Having a panoramic view also means being able to gather a wealth of information from other foragers, be they conspecifics or otherwise \citep{jackson2008effect}.
Again, returning to vultures, the genus \textit{Gyps} consists of highly social and colonially nesting species \citep{fernandez2015density}.
These behaviours allow them forage far more efficiently because one bird can scrounge information on the location of food from another successful forager \citep{KaneVul}.

@@

\section*{Handling Time}
Since carrion is not dispatched directly, often the most easily accessible and choicest components of the carcass will be missing or, if present, will be fought over.
Being able to extract nutrients from remnants gives a scavenger a great advantage.
%Thus, the bone crushing ability of hyenas and others reveals another useful scavenger trait. %TG: how about the following for a smoother transition:
Thus, some phenotypic adaptations such as the ability to crush bone (as seen in hyenas) reveals another useful scavenger trait.
Osteophagy is known across a range of terrestrial carnivores and given that some fat-rich mammalian bones have an energy density (6.7 kJ/g) comparable with that of muscle tissue, it makes skeletal remains an enticing resource \citep{brown1989study}.
This ability reached its zenith among hyenas with the evolution of the 110 kg \textit{Pachycrocuta brevirostris} during the Pliocene \citep[3.6 - 2.58 Mya; ][]{palmqvist2011giant}.
Indeed, their extinction has been blamed on the decline of sabretooth cats (Machairodontinae), the unique skull morphology of the latter meant they would leave a large amount of food on a carcass for would-be scavengers \citep{palmqvist2011giant}. 
Earlier in the evolution of mammals, the bone-crushing dogs that evolved during the Oligocene (Borophaginae; 33.9 - 23.03 Mya) have also been compared to hyenas in terms of their feeding ecology \citep{van2003chapter,martin2016pursuit}.

% Some work on extinct sabretooths suggests they may have left a large amount of food for would-be scavengers because of their unique skull morphology.
% As a result, the decline of Machairodontinae sabretooths has been offered as an explanation for the extinction of \textit{P brevirostris} \citep{palmqvist2011giant}.
% And many of the aforesaid adapations for scavenging are found in these other major terrestrial mammalian carnviores.
% Though the specific mix of features realised in hyenas suggest this is the model organism for terrestrial scavenging among mammals in the past.
% These traits have been used to infer scavenging lifestyles in extinct mammals.

@@

Large body size confers substantial dominance and starvation-resistance benefits \citep{ruxton2004obligate}.
Thus, we would expect scavengers to have this trait selected for even in the case of weight-constrained fliers.
Wandering Albatrosses (\textit{Diomedea exulans}), cinereous vultures (\textit{Aegypius monachus}) and condors (\textit{Vultur gryphus}, \textit{Gymnogyps californianus}) all have body masses that can exceed 10 kg and represent some of the heaviest bird species capable of flight \citep{weimerskirch1992reproductive,ferguson2001raptors,donazar2002effects}.
And as we have noted the Azhdarchid pterosaurs were far bigger again, with estimated body masses of over 200 kg \citep{witton2010size}.
Although \cite{witton2008reappraisal} argued that neck inflexibility and straight, rather than hooked jaw morphology points against pterosaurs existing as \textit{obligate} scavengers, Azhdarchid terrestrial proficiency indicates they would have been comfortable foraging on the ground.
Indeed, extant Marabou Storks (\textit{Leptoptilos crumenifer}) have a comparable morphology and are noted facultative scavengers \citep{monadjem2012survival} so it is reasonable to believe that certain pterosaurs behaved similarly.

%%%%%%%%%%  AK This annoying paragraph's new home
Certainly, scavenging should be particularly attractive to flying species compared to mammals. % Not sure about this paragraph 
The latter can kill prey up to the same body mass as themselves and sometimes an order of magnitude heavier \citep[e.g. socially hunting lions; ][]{owen2008predator}.
In contrast, birds of prey tend to kill prey smaller than themselves \citep{slagsvold2007prey} because of the greater cost of injury and the need to carry off their food.
% This is likely due to their need to kill animals that they can fly away with, as well as the risk of injury being higher (which carries a higher mortality risk) for a bird than a mammal.
Scavenging provides a means for birds to exploit species that would otherwise be too big for them to kill.
% Thus, for birds, scavenging means they can exploit species that would otherwise be too big for them to kill.
%%%%%%%%%%this is moved from the locomotion section as it didn't seem to fit.

%TG: This part above makes sense in the adaptations parts (i.e. in the Handling Time part) but we can see when we clean down the whole paper if we keep it (then we need to integrate it) or if we have enough. I actually fits way better in the new narrative by the way!
On the ground, the competitive ability of even the largest flying bird is radically diminished in their interactions with mammalian competitors however, and as such they tend to consume carrion rapidly. %TG: I've added "living" bird here. In the Paleocene you have these giant terror birds that were still top predators (like during the Cretaceous... How weird! ;)). 
% AK How about flying bird because that's the key thing here 
\cite{houston1974role} observed a group of \textit{Gyps} vultures consuming all of the soft tissue from a 50 kg Grant’s gazelle (\textit{Nanger granti}) in eight minutes. 
Their serrated tongues and hooked bills enabling them to achieve this feat \citep{houston1975digestive}. 
Outside of raptors such as vultures, the specialised beaks of many modern bird lineages hinders their ability to eat meat which is in contrast to the first lineages that did not have this feature \citep{martyniuk2012field}. 
As \cite{martyniuk2012field} notes these early birds would thus have been predominantly carnivorous, which implies that scavenging would have been a live opportunity cf. their descendants.
